---
title: "Session 4 - Solutions"
author: "L. Wegner"
format: html
theme: minty
editor: visual
---

# Today's Exercise

## Task 1 - Using functions

Take this vector:

``` r
vector <- c(19,3,8,7,24,43,89,76,17,32,8,9,350,56)
```

1\. What is the lowest number of the vector?

```{r} vector <- c(19,3,8,7,24,43,89,76,17,32,8,9,350,56) min(vector)}
```

2\. What is the mean value? What is the median value? How far are they apart from each other?

```{r} vector <- c(19,3,8,7,24,43,89,76,17,32,8,9,350,56) mean <- mean(vector) median <- median(vector) diff <- mean - median diff}
```

3\. How big is the range of the vector?

```{r} vector <- c(19,3,8,7,24,43,89,76,17,32,8,9,350,56) range(vector)}
```

## Task 2 - Calculator Function

-   We want a function that calculates the sum of the square roots of two numbers,

    ```{r} calcu <- function(x,y) {  root_x <- sqrt(x)  root_y <- sqrt(y)   result <- root_x + root_y   return (result) }
    ```

-   rounds it to two digits

    ```{r} calcu <- function(x,y) {   root_x <- sqrt(x)   root_y <- sqrt(y)   result <- root_x + root_y   rounded <- round(result, digits =2)   return (round) }
    ```

-   and then does a check: when the result is bigger then a third number return: bigger then z, otherwise return: smaller then z

```{r} calcu <- function(x, y, z) {   root_x <- sqrt(x)   root_y <- sqrt(y)   result <- root_x + root_y   rounded <- round(result, digits = 2)   if (rounded >= z) {     return("result is bigger then z")   } else {     return("result is smaller then z")   } }
```

-   check by using x=10.33, y=20.78 and z=40.47

    ```{r} calcu(10.33,20.78,40.47)}
    ```

## 3. loading data into R

| 1. first load either the data-set "Exam_Score" or "California Housing" (can be found at OpenOlat) into R using package-functions of readr (the data-sets are both in csv-format).
| Describe the steps of your process via comments.

Installing the "Readr" package

``` r
#install.packages("readr")
library(readr)
```

Loading the csv-file from either the internet

``` r
cal_housing_internet <- read_csv("Link_that_ends_with_csv.csv")
exam_data_internet <- read_csv("link_that_also_ends_with_csv.csv")
```

.... or from it's folder on my computer

I download the file from OpenOlat, then store it in a folder and give this path to the function.

::: callout-note
Note: .csv needs to be the ending of my file-name here as well!
:::

::: callout-caution
The file-path on my computer is: Z:\\\\wiso\\Datasets\\cal_housing.csv

**R can not take this! I have to change the \\ to /**

We can do that with Strg + F and then put \\ in the **find-field** and **replace** it by /

Z://wiso/Datasets/cal_housing.csv
:::

``` r
cal_housing_laptop <- read_csv("Z://Datasets/edited_California_Houses.csv")

exam_data_laptop <- read_csv("Z://Datasets/exam score/Expanded_data_studentexams_soz.csv")
```

| 2. load the other data-set from our package "sozoekds" (load & install this package first, then load the data-set)

Install devtools and load to library if you have not done that yet (remove the \# from the code)

```{r}
#install.packages("devtools")
#library(devtools)
#devtools::install_git("https://gitlab.rrz.uni-hamburg.de/baq6370/sozoekds.git")
library(sozoekds)
```

Load the data set from the package - under "packages" you can click on sozoekds and find out the names of the data sets

```{r}
# first there is the new name (left side) - on the right side there is the name of the data set in the package
cal_housing <- calhouse
exam_data <- examscores

```

## 4. descriptive analysis

<div>

| How many variables are there?

</div>

exam_data includes 15 variables

| Type of those variables?

I can find out the type of a variable three ways (works for both data sets the same way)

```{r}
class(exam_data$MathScore)
```

Or by checking "Data" in the "Environment" on the right side - I can see the name of the data set, its variables and their types and get a glimpse of the values.

If I click on the little table (right upper corner of picture) I can see a full table - by hovering over the variable names I can find out their types as well.

# Exam Scores

### descriptive analysis

| Calculations: Mean, Median, Range - what do you think is helpful (and why) to get a good overview over your data

```{r}
# for the writing score a mean value would be helpful
mean(exam_data$WritingScore)
# Interpretation: the mean writing score is 68,4 points - the contestants had on average almost 70% of the writing tasks correct.
```

```{r}
# range of Math Scores shows the highest and lowest scores in this category.
range(exam_data$MathScore)
# Interpretation: One (or more) persons hat zero points, one or more had 100 points

```

| Are there other ways to get an overview over the data?

```{r}
# How many persons had zero points?
# The == is a check if the Score equals the 0 - persons where this expression is true get summed up
sum(exam_data$MathScore == 0)
# Interpretation: 1 Person had 0 points

# How many had 100 points?
sum(exam_data$MathScore == 100)
# Interpretation: 275 persons had everything right.
```

Categorical variables can also be summed up

```{r}
# Table of the Weekly Study Hours 
result <- table(exam_data$WklyStudyHours) 
print(result) 
```

Interpretation: for 955 persons there is no answer, 8238 persons studied under 5h per week.... etc.}

### Transformation

| Transform ethnic group to factor

```{r}
exam_data$EthnicGroup <- as.factor(exam_data$EthnicGroup)

str(exam_data$EthnicGroup)
```

There are 6 ethnic groups within the data set

| calculate the total points in all tests per person
| use 300 as 100% and calculate a percentage using the total points

### total points

```{r}
for (i in exam_data) { exam_data$total_score <- exam_data$WritingScore + exam_data$MathScore + exam_data$ReadingScore }

for (i in exam_data) { exam_data$grade_percent <- round(exam_data$total_score/(300/100), digits = 0) }
```

### passed?

with more than 50% a person has passed

```{r}
exam_data$passing <- ifelse(exam_data$grade_percent < 50, "TRUE", "FALSE")
```
